---
title: "CC EDA"
author: "Chris Joyce"
date: "2025"
output: 
  html_document:
    number_sections: false
    toc: true
    toc_depth: 2
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyverse)
library(skimr)
library(tableone)
library(forecast)
library(tseries)
library(astsa)
library(xts)
library(C50)
library(tidymodels)
library(xgboost)
library(caret)
library(future)
library(lubridate)
plan(multisession, workers = 6)
```

# Objective

### The objective of this notebook is to provide exploratory insights into the historical data pertaining to Swire Coca Cola.  The contents consist of preliminary data preparation steps and transformations intended to provide usable data for analytical usage, and discover patterns or attributes of interest within the data to aid in solving the presented business problem.  Descriptions of steps and deductive statements are provided as needed.


# Load in and Prep


```{r, warning=FALSE}
setwd("F:/Lori/Downloads")
td <- read.csv('transactional_data.csv')
cust <- read.csv('customer_profile.csv')
cust_a <- read.csv('customer_address_and_zip_mapping.csv')
cost <- read.csv('delivery_cost_data.csv')
m <- read.csv('merged_data.csv')
```

```{r}
str(td)
summary(td)
head(td, 50)
td$ORDER_TYPE <- td$ORDER_TYPE %>% factor()
```

```{r}
skim(td)
```

```{r}
# Unique Customers
cust_chrs <- cust[, c(3,6,7,8)]
cust_chrs <- lapply(cust_chrs, as.factor)
cust_chrs <- as.data.frame(cust_chrs)
cust <- cbind(cust[, -c(3,6,7,8)], cust_chrs)
```

```{r}
cust <- cust %>% rename(zip = ZIP_CODE)
str(cust)
```

```{r}
skim(cust)
```

```{r}
# Unique Zips
str(cust_a)
```

```{r}
cust_a
```

```{r}
custs_match <- merge(cust, cust_a, by = "zip")
```

```{r}
str(custs_match) # customers with full address info. no rows lost
```

```{r}
skim(custs_match)
```

```{r}
# 12,282 belong to PG
custs_match |>
  group_by(PRIMARY_GROUP_NUMBER)|>
  arrange(PRIMARY_GROUP_NUMBER) |>
  na.omit()
```

```{r}
cost
```

```{r}
cost$Median.Delivery.Cost <- gsub("\\$", "", cost$Median.Delivery.Cost)
cost$Median.Delivery.Cost <- as.numeric(cost$Median.Delivery.Cost)
```

```{r}
split_vol_range <- function(range) {
  if (range == "1350+") {
    return(c(1350, NA))
  } else {
    parts <- strsplit(range, " - ")[[1]]
    return(as.numeric(parts))
  }
}
```

```{r}
vol_ranges <- t(sapply(cost$Vol.Range, split_vol_range))
cost$Vol.Range.Min <- vol_ranges[, 1]
cost$Vol.Range.Max <- vol_ranges[, 2]
cost <- cost |>
  mutate(Vol.Range.Max = if_else(Vol.Range.Min == 1350, 1350, Vol.Range.Max))
```

```{r}
cost <- cost |>
  rename(COLD_DRINK_CHANNEL = Cold.Drink.Channel)
```

```{r}
cost
```

```{r}
# Matching orders to customers
cust_tot <- merge(custs_match, td, on = "CUSTOMER_NUMBER")
```

```{r}
# Removing negative order volumes
cust_tot <- cust_tot |>
  mutate(DELIVERED_CASES = ifelse(DELIVERED_CASES < 0, 0, DELIVERED_CASES),
         ORDERED_CASES = ifelse(ORDERED_CASES < 0, 0, ORDERED_CASES),
         LOADED_CASES = ifelse(LOADED_CASES < 0, 0, LOADED_CASES),
         ORDERED_GALLONS = ifelse(ORDERED_GALLONS < 0, 0, ORDERED_GALLONS),
         LOADED_GALLONS = ifelse(LOADED_GALLONS < 0, 0, LOADED_GALLONS),
         DELIVERED_GALLONS = ifelse(DELIVERED_GALLONS < 0, 0, DELIVERED_GALLONS))
```

```{r}
#cust_combi <- merge(cust_tot, cost, on = "COLD_DRINK_CHANNEL")
```

```{r}
str(cust_tot)
```

```{r}
summary(cust_tot)
```

```{r}
skim(cust_tot)
```

```{r}
head(cust_tot)
```


# EDA


```{r}
ggplot(data = cust_tot, 
       aes(x = ORDER_TYPE)) + 
  geom_bar() +
  labs(title = "Count of Order Types")
```

```{r}
distribution <- table(cust_tot$FREQUENT_ORDER_TYPE)

prop.table(distribution) * 100
```

```{r}
distribution <- table(cust_tot$TRADE_CHANNEL)

prop.table(distribution) * 100
```

```{r}
distribution <- table(cust_tot$SUB_TRADE_CHANNEL)

prop.table(distribution) * 100
```

```{r}
# Distribution of cold drink channel orders
ggplot(data = cust_tot, 
       aes(x = COLD_DRINK_CHANNEL)) + 
  geom_bar() +
  labs(title = "Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_x_discrete(drop = FALSE) 
```

```{r}
# Verify low conventional count
cust_tot |>
  filter(COLD_DRINK_CHANNEL == "CONVENTIONAL")
cust_tot |>
  filter(COLD_DRINK_CHANNEL == "CONVENTIONAL") |>
  count(CUSTOMER_NUMBER) |>
  arrange(desc(n))
```

```{r}
# Trade Channel Frequency of Orders
cust_tot |>
  group_by(TRADE_CHANNEL) |>
  count() |>
  arrange(desc(n))
```

```{r}
# Sub Trade Channel Frequency of Orders
cust_tot |>
  group_by(SUB_TRADE_CHANNEL) |>
  count() |>
  arrange(desc(n))
```

```{r}
# Frequent Order Type Frequency of Orders
cust_tot |>
  group_by(FREQUENT_ORDER_TYPE) |>
  count() |>
  arrange(desc(n))
```

```{r}
# Composition of trade channels
TRADE_CHANNELS <- unique(cust_tot$TRADE_CHANNEL)

for (channel in TRADE_CHANNELS) {
  df <- cust_tot |>
    select(TRADE_CHANNEL, SUB_TRADE_CHANNEL) |>
    filter(TRADE_CHANNEL == channel) |>
    distinct(SUB_TRADE_CHANNEL) |>
    mutate(TRADE_CHANNEL = channel)
  
  assign(paste0("df_", gsub(" ", "_", channel)), df)
  print(df)
}
```

```{r}
# Distribution of Customers Belonging to a Primary Group
cust_tot$PRIMARY_GROUP_STATUS <- ifelse(is.na(cust_tot$PRIMARY_GROUP_NUMBER), "NA", "Not NA")
vars = c("LOCAL_MARKET_PARTNER", "CO2_CUSTOMER")
groups = "PRIMARY_GROUP_STATUS"
pgroup_table <- CreateTableOne(vars = vars, strata = groups, data = cust_tot)
pgroup_table
```

```{r}
# Creating bins for single order volumes
cust_tot_mod <- cust_tot |>
  filter(ORDERED_CASES <= 10 & ORDERED_CASES > 0, ORDERED_GALLONS <= 10 & ORDERED_GALLONS > 0)
cust_tot_mod2 <- cust_tot |>
  filter(ORDERED_CASES > 10 & ORDERED_CASES <= 50, ORDERED_GALLONS > 10 & ORDERED_GALLONS <= 50)
cust_tot_mod3 <- cust_tot |>
  filter(ORDERED_CASES > 50, ORDERED_GALLONS > 50)
```

```{r}
cust_tot_long <- cust_tot_mod %>%
  pivot_longer(cols = c(ORDERED_CASES, ORDERED_GALLONS), names_to = "Order_Type", values_to = "Value")
cust_tot_long2 <- cust_tot_mod2 %>%
  pivot_longer(cols = c(ORDERED_CASES, ORDERED_GALLONS), names_to = "Order_Type", values_to = "Value")
cust_tot_long3 <- cust_tot_mod3 %>%
  pivot_longer(cols = c(ORDERED_CASES, ORDERED_GALLONS), names_to = "Order_Type", values_to = "Value")
```

```{r}
ggplot(data = cust_tot_long, aes(x = Value, fill = Order_Type)) +
  geom_histogram(bins = 5, position = "dodge") +
  facet_wrap(~ Order_Type) +
  labs(title = "Histogram for Small Orders",
       x = "Order Amount",
       y = "Count") +
  theme_minimal()
```

```{r}
ggplot(data = cust_tot_long2, aes(x = Value, fill = Order_Type)) +
  geom_histogram(bins = 5, position = "dodge") +
  facet_wrap(~ Order_Type) +
  labs(title = "Histogram for Medium Orders",
       x = "Order Amount",
       y = "Count") +
  theme_minimal()
```

```{r}
ggplot(data = cust_tot_long3, aes(x = Value, fill = Order_Type)) +
  geom_histogram(position = "dodge", binwidth = 500, bins = 5) +
  facet_wrap(~ Order_Type) +
  labs(title = "Histogram for Large Orders",
       x = "Order Amount",
       y = "Count") +
  theme_minimal()
```

```{r}
# Average orders for different trade channels
order_dists_all <- cust_tot |>
  group_by(TRADE_CHANNEL) |>
  summarize(gallon_avg = mean(ORDERED_GALLONS), case_avg = mean(ORDERED_CASES), total_avg = case_avg + gallon_avg,
            case_ratio = case_avg / total_avg, gallon_ratio = gallon_avg / total_avg) |>
  arrange(desc(total_avg))
order_dists_all
```

```{r}
# Number of customers belonging to each primary
cust_tot %>%
  group_by(PRIMARY_GROUP_NUMBER) %>%
  summarise(count = n_distinct(CUSTOMER_NUMBER),
            case_avg = mean(ORDERED_CASES),
            gal_avg = mean(ORDERED_GALLONS),
            trade_channels = paste(unique(TRADE_CHANNEL),
            collapse = ", ")) %>%
  arrange(desc(count))
```

```{r}
cust_tot |>
  select(TRADE_CHANNEL, COLD_DRINK_CHANNEL) |>
  filter(COLD_DRINK_CHANNEL == "CONVENTIONAL")
```

```{r}
# Type casting
m_chrs <- m[, c(5, 13, 16, 17, 18)]
m_chrs <- lapply(m_chrs, as.factor)
m_chrs <- as.data.frame(m_chrs)

m[, c(5, 13, 16, 17, 18)] <- m_chrs

m_log <- m[, c(19, 20)]
m_log <- lapply(m_log, as.logical)
m_log <- as.data.frame(m_log)

m[, c(19, 20)] <- m_log
```

```{r}
str(m)
```

```{r}
# LMP analysis
LMP_distinct <- m |>
  filter(LOCAL_MARKET_PARTNER == TRUE, CO2_CUSTOMER == FALSE, ORDERED_CASES == 0, LOADED_CASES == 0, DELIVERED_CASES == 0)
  #summarize(case_avg = mean(ORDERED_CASES), gallon_avg = mean(ORDERED_GALLONS), total_avg = case_avg + gallon_avg) |>
  #arrange(case_avg)
LMP_distinct
```

### There are 64,135 orders for local market partners who exclusively order gallons.  This is one of the distinct customer groups of interest for the business problem.

```{r}
# Trade channel averages for LMP customers
order_dists_lmp <- LMP_distinct |>
  group_by(TRADE_CHANNEL) |>
  summarize(gallon_avg = mean(ORDERED_GALLONS)) |>
  arrange(desc(gallon_avg))
order_dists_lmp
```

```{r}
str(m)
```

```{r}
# Individual customer orders for each year in LMP group
g1 <- LMP_distinct |>
  select(CUSTOMER_NUMBER, zip, PRIMARY_GROUP_NUMBER, LOCAL_MARKET_PARTNER, FREQUENT_ORDER_TYPE, COLD_DRINK_CHANNEL,
         TRADE_CHANNEL, SUB_TRADE_CHANNEL, TRANSACTION_DATE, ORDERED_CASES, ORDERED_GALLONS, Month, Has.Primary.Group.Number., YEAR) |>
  group_by(CUSTOMER_NUMBER, YEAR) |>
  summarise(TOTAL_YEAR_ORDERS = sum(ORDERED_CASES + ORDERED_GALLONS, na.rm = TRUE),
            zip = first(zip),
            PRIMARY_GROUP_NUMBER = first(PRIMARY_GROUP_NUMBER),
            LOCAL_MARKET_PARTNER = first(LOCAL_MARKET_PARTNER),
            FREQUENT_ORDER_TYPE = first(FREQUENT_ORDER_TYPE),
            COLD_DRINK_CHANNEL = first(COLD_DRINK_CHANNEL),
            TRADE_CHANNEL = first(TRADE_CHANNEL),
            SUB_TRADE_CHANNEL = first(SUB_TRADE_CHANNEL),
            TRANSACTION_DATE = first(TRANSACTION_DATE),
            Month = first(Month),
            Has.Primary.Group.Number. = first(Has.Primary.Group.Number.),
            .groups = 'drop') |>
  ungroup()

g1
```

```{r}
# Individual customer orders for each year (all customers)
g2 <- m |>
  select(CUSTOMER_NUMBER, zip, PRIMARY_GROUP_NUMBER, LOCAL_MARKET_PARTNER, CO2_CUSTOMER, FREQUENT_ORDER_TYPE, COLD_DRINK_CHANNEL,
         TRADE_CHANNEL, SUB_TRADE_CHANNEL, TRANSACTION_DATE, ORDERED_CASES, ORDERED_GALLONS, Month, Has.Primary.Group.Number., YEAR) |>
  group_by(CUSTOMER_NUMBER, YEAR) |>
  summarise(TOTAL_YEAR_ORDERS = sum(ORDERED_CASES + ORDERED_GALLONS, na.rm = TRUE),
            zip = first(zip),
            PRIMARY_GROUP_NUMBER = first(PRIMARY_GROUP_NUMBER),
            LOCAL_MARKET_PARTNER = first(LOCAL_MARKET_PARTNER),
            CO2_CUSTOMER = first(CO2_CUSTOMER),
            FREQUENT_ORDER_TYPE = first(FREQUENT_ORDER_TYPE),
            COLD_DRINK_CHANNEL = first(COLD_DRINK_CHANNEL),
            TRADE_CHANNEL = first(TRADE_CHANNEL),
            SUB_TRADE_CHANNEL = first(SUB_TRADE_CHANNEL),
            TRANSACTION_DATE = first(TRANSACTION_DATE),
            Month = first(Month),
            Has.Primary.Group.Number. = first(Has.Primary.Group.Number.),
            .groups = 'drop') |>
  ungroup()

g2
```






